---
title: "Session02"
author: "Susumu Shikano"
date: "Last compiled at `r format(Sys.Date(), '%d. %B %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Applying Bayes' rule to our example 


Here, we first begin to obtain the posterior probability after observing 60 out of 100 trials is successful:

We can first calculate $P(D|H_A)$ and $P(D|H_B)$:

```{r }
print(prob.D.HA <- choose(100,60) * (0.7^60)*(0.3^40))

print(prob.D.HB <- choose(100,60) * ((0.5)^100))

```

Let's specify prior:

```{r}
prob.HA <- 0.5

prob.HB <- 0.5

```


Now, we can apply the Bayes' rule:

```{r}

prob.D.HA*prob.HA /(prob.D.HA*prob.HA+prob.D.HB*prob.HB)

prob.D.HB*prob.HB /(prob.D.HA*prob.HA+prob.D.HB*prob.HB)

```


### Sequential updating by using the Bayes' rule

Now, we can compute posterior sequentially. That is, we update posterior after each trial and use it as prior for the next trial.

To do this, we first define a function:

```{r}

bayes.update <- function(prior,success,H){
  enumerator.1  <- H[1]^success * (1-H[1])^(1-success) * prior[1]
  enumerator.2  <- H[2]^success * (1-H[2])^(1-success) * prior[2]
  
  posterior.1 <- enumerator.1/(enumerator.1 + enumerator.2)
  posterior.2 <- enumerator.2/(enumerator.1 + enumerator.2)
  
  c(posterior.1,posterior.2)
}

```


This function requires three inputs:

* prior: prior vector for $H_A$ and $H_B$.
* success: a dummy variable for successful judgement.
* H: predicted success rate for $H_A$ and $H_B$.

We can just test by using an example:

```{r}
prior <- c(0.5,0.5)
success <- 1
H <- c(0.7,0.5)


print(updated <- bayes.update(prior,success,H))
```

Now, we can use the obtained posterior as prior for the next trial:

```{r}
bayes.update(updated,success,H)
```

Now let's create a permutation of 100 trials in which 60 are successful:

```{r}

all.trials <- sample(c(rep(1,60) , rep(0,40)),
                replace=F)
print(all.trials)
table(all.trials)

```

By using this data and the function above, we can sequentially update posterior:

```{r}
first.prior <- c(0.5,0.5)
all.updates <- matrix(NA,nrow=length(all.trials),ncol=2)

for (i in 1:length(all.trials)){
  
  if (i ==1) updated <- first.prior
  
  updated <- bayes.update(updated,all.trials[i],H)
  
  all.updates[i,] <- updated
}
```


We visualize the update process:

```{r}


plot(1:nrow(all.updates),all.trials,type="p",col="grey",
     ylab="Posterior",xlab="Trials",axes=F,
     ylim=c(0,1),
     xlim=c(0,length(all.trials)*1.3))
axis(1,at=seq(0,100,by=20))
axis(2)
par(new=T)
plot(1:nrow(all.updates),all.updates[,1],type="l",col="red",
     ann=F,axes=F,
     ylim=c(0,1),
     xlim=c(0,length(all.trials)*1.3))
par(new=T)
plot(1:nrow(all.updates),all.updates[,2],type="l",col="blue",
     ann=F,axes=F,
     ylim=c(0,1),
     xlim=c(0,length(all.trials)*1.3))
legend("topright",col=c("grey","red","blue"),pch=c(1,NA,NA),lty=c(NA,1,1),
       c("Success","H_A","H_B"),bty="n")



```





